---
title: "Basic workflow: from cell phase inference to rhythm-state modules"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Basic workflow: from cell phase inference to rhythm-state modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  fig.retina = 2,
  out.width = "85%"
)
```

# Overview

`scRhythmBayes` is designed for phase-resolved circadian analysis of heterogeneous single-cell transcriptomic data.

This tutorial demonstrates the basic workflow using the built-in example dataset. The workflow follows three core analysis layers:

1. **Cell-level circadian phase inference**  
   Infer a latent circadian phase for each cell.

2. **Gene-level rhythmic parameter inference**  
   Estimate rhythmic parameters, including mesor, mean expression, amplitude, phase, dropout rate, likelihood-ratio statistics, and FDR.

3. **Differential rhythm-state module analysis**  
   Classify genes into interpretable rhythm-state remodeling modules between two biological groups.

The main example compares condition `A` versus condition `B` within one cell type. This is the recommended starting point for two-condition rhythm-state remodeling analysis. The tutorial also shows how the same module framework can compare two cell types when gene-level parameters are estimated by `celltype`.

Extension analyses, including one-vs-rest rhythmic program discovery, DTW-based phase-distribution similarity, and circular phase-concentration rho analysis, are introduced briefly at the end and described in separate tutorials.

# Load package

```{r load-package}
library(scRhythmBayes)
```

# Load example data

`scRhythmBayes` includes a simulated example dataset for testing the complete workflow.

```{r load-data}
data("srb_example_data")

names(srb_example_data)
```

The example object contains a gene-by-cell count matrix and cell-level metadata.

```{r inspect-data}
dim(srb_example_data$Count)
class(srb_example_data$Count)

dim(srb_example_data$meta)
head(srb_example_data$meta)
```

The two main input objects are:

| Object | Description |
|---|---|
| `srb_example_data$Count` | Gene-by-cell sparse count matrix |
| `srb_example_data$meta` | Cell-level metadata |

The metadata contains:

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT` | Sampling Zeitgeber time |
| `celltype` | Cell-type label |
| `condition` | Condition or group label |
| `T_true` | Simulated true circadian time in hours |
| `Theta_true` | Simulated true circadian phase in radians |

`T_true` and `Theta_true` are available only in the simulated example dataset. Real single-cell datasets usually do not contain these two columns.

For convenience, define the input count matrix and metadata:

```{r define-inputs}
count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

# Choosing the analysis design

Before running gene-level rhythmic parameter inference, users should define the biological comparison of interest. In `scRhythmBayes`, the same core framework can support several common designs.

| Analysis goal | Recommended design | Parameter grouping | Downstream analysis |
|---|---|---|---|
| Compare two conditions within one cell type | Subset one cell type, then compare conditions | `strata_by = "condition"` | `infer_rhythm_module(group1 = "A", group2 = "B")` |
| Compare two cell types directly | Use all relevant cells, then estimate parameters by cell type | `strata_by = "celltype"` | `infer_rhythm_module(group1 = "1", group2 = "2")` |
| Identify cell-type-specific rhythmic programs | Estimate parameters by cell type | `strata_by = "celltype"` | `infer_ovr_rhythm()` |
| Compare phase distributions across cell types and conditions | Use cell-level phase result and metadata | not required | `run_dtw_similarity()` or `run_phase_rho()` |

The recommended starting point is a two-condition comparison within one biologically defined cell type. This design is usually the clearest setting for interpreting differential rhythm-state modules.

A direct cell-type comparison and a one-vs-rest analysis answer different questions. A direct module comparison tests one selected pair, such as cell type `1` versus cell type `2`. A one-vs-rest analysis identifies target-specific rhythmic programs by comparing each target cell type against all remaining cell types.

# Step 1: infer cell-level circadian phase

## Input

The phase inference step requires:

| Argument | Description |
|---|---|
| `count` | Gene-by-cell count matrix |
| `metadata` | Cell-level metadata |
| `ZT_by` | Column containing sampling Zeitgeber time |
| `strata_by` | Column defining strata for phase inference, such as cell type |
| `Species` | Species used for selecting default seed rhythmic genes |

## Run phase inference

```{r infer-cell-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

## Output

The output contains one row per cell.

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT_input` | Observed sampling Zeitgeber time |
| `stratum` | Stratum used during phase inference |
| `theta_pred` | Inferred cell-level circadian phase in radians |
| `ZT_pred` | Inferred cell-level circadian phase converted to 24-hour time |
| `Rbar` | Aggregate phase concentration statistic |
| `r_cell` | Cell-level phase reliability weight |

The inferred cell phase is used as the latent temporal coordinate for downstream gene-level rhythmic parameter estimation.

## Phase diagnostic plots

If simulated true time is available, inferred phase can be compared against the known true time.

```{r phase-scatter, fig.width=6.5, fig.height=5, out.width="80%"}
p_phase <- plot_phase_scatter(
  res_phase = res_phase,
  metadata = metadata,
  truth_by = "T_true",
  pred_by = "ZT_pred"
)

p_phase$plot
```

For real datasets, true phase is usually unknown. In that case, circular phase distributions can be inspected by condition or cell type.

```{r phase-circle-condition, fig.width=6.5, fig.height=5, out.width="80%"}
p_circle_condition <- plot_phase_circle(
  res_phase = res_phase,
  metadata = metadata,
  group_by = "condition",
  theta_by = "theta_pred"
)

p_circle_condition$plot
```

```{r phase-circle-celltype, fig.width=6.5, fig.height=5, out.width="80%"}
p_circle_celltype <- plot_phase_circle(
  res_phase = res_phase,
  metadata = metadata,
  group_by = "celltype",
  theta_by = "theta_pred"
)

p_circle_celltype$plot
```

# Step 2: select one cell type for two-condition comparison

The core module workflow is designed for a defined two-group comparison. In this example, we compare condition `A` and condition `B` within cell type `1`.

```{r select-celltype}
cell_use <- as.character(metadata$celltype) == "1"

count_use <- count[, cell_use, drop = FALSE]
metadata_use <- metadata[cell_use, , drop = FALSE]
res_phase_use <- res_phase[cell_use, , drop = FALSE]

dim(count_use)
table(metadata_use$condition)
```

This design mirrors a typical biological analysis where rhythmic remodeling is tested within a defined cell type or tissue compartment.

# Step 3: estimate gene-level rhythmic parameters

## Input

The gene-level model requires:

| Argument | Description |
|---|---|
| `count` | Gene-by-cell count matrix for the selected analysis subset |
| `metadata` | Metadata for the selected cells |
| `theta_pred` | Inferred cell-level phase |
| `r_cell` | Cell-level phase reliability weight |
| `strata_by` | Grouping variable for parameter estimation |

For condition-level module analysis within one cell type, use `strata_by = "condition"`. For cell-type-level comparison, use `strata_by = "celltype"` on the full dataset or a larger subset.

## Run condition-level rhythmic parameter inference

```{r infer-gene-param-condition}
res_param <- infer_gene_param(
  count = count_use,
  metadata = metadata_use,
  theta_pred = res_phase_use$theta_pred,
  r_cell = res_phase_use$r_cell,
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

## Optional: run cell-type-level rhythmic parameter inference

This second parameter table is used only to demonstrate a direct cell type `1` versus cell type `2` module comparison.

```{r infer-gene-param-celltype}
res_param_celltype <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

## Output

The result contains long-format and wide-format gene-level parameter tables.

```{r show-param-long}
head(res_param$param_long)
head(res_param_celltype$param_long)
```

```{r show-param-wide}
head(res_param$param_wide)
head(res_param_celltype$param_wide)
```

Main output objects:

| Object | Description |
|---|---|
| `param_long` | Long-format gene-level rhythmic parameter estimates for each group |
| `param_wide` | Wide-format paired parameter table for downstream module analysis |
| `bootstrap_res` | Bootstrap confidence interval results when available |

Common gene-level parameters:

| Column | Description |
|---|---|
| `gene` | Gene name |
| `Mesor` | Baseline rhythmic expression level |
| `Mean` | Mean expression level |
| `Amplitude` | Circadian oscillation amplitude |
| `Phase` | Gene-level rhythmic phase in radians |
| `Alpha` | Dispersion-related parameter |
| `Dropout_rate` | Estimated dropout rate |
| `LRT_stat` | Likelihood-ratio test statistic |
| `P_value` | Nominal rhythmicity test P value |
| `FDR` | Multiple-testing adjusted rhythmicity significance |

# Step 4: infer differential rhythm-state modules

The rhythm-state module analysis compares rhythmic parameters between two groups and assigns genes to interpretable remodeling classes.

## Input

| Argument | Description |
|---|---|
| `param_wide` | Wide-format paired gene parameter table from `infer_gene_param()` |
| `compare_by` | Name of the comparison variable |
| `group1`, `group2` | Two groups to compare |
| `out_dir` | Optional output directory |

The `param_wide` object must match the selected groups. For example, comparing `group1 = "A"` and `group2 = "B"` requires columns such as `Mean_A`, `Mean_B`, `FDR_A`, `FDR_B`, `Amplitude_A`, `Amplitude_B`, `Phase_A`, and `Phase_B`. Comparing `group1 = "1"` and `group2 = "2"` requires columns such as `Mean_1`, `Mean_2`, `FDR_1`, `FDR_2`, `Amplitude_1`, `Amplitude_2`, `Phase_1`, and `Phase_2`.

## Run condition-level module inference

```{r infer-module-condition}
res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

head(res_module$module_result)
```

## Optional: run direct cell-type-level module inference

This optional example compares cell type `1` and cell type `2` directly. It uses `res_param_celltype$param_wide`, not `res_param$param_wide`, because the wide-format columns must match the selected group labels.

```{r infer-module-celltype}
res_module_celltype <- infer_rhythm_module(
  param_wide = res_param_celltype$param_wide,
  compare_by = "celltype",
  group1 = "1",
  group2 = "2",
  out_dir = NULL,
  Verbose = FALSE
)

head(res_module_celltype$module_result)
```

## Output

The module result contains gene-level comparison statistics and posterior module assignments.

| Column | Description |
|---|---|
| `gene` | Gene name |
| `group1`, `group2` | Compared groups |
| `FDR_A`, `FDR_B` or `FDR_1`, `FDR_2` | Rhythmicity FDR in each group |
| `Amplitude_A`, `Amplitude_B` or `Amplitude_1`, `Amplitude_2` | Rhythmic amplitude in each group |
| `Phase_A`, `Phase_B` or `Phase_1`, `Phase_2` | Rhythmic phase in each group |
| `delta_phi_rad` | Phase shift between groups in radians |
| `module_name` | Assigned rhythm-state remodeling module |
| `post_max` | Maximum posterior module probability |
| `margin` | Difference between top posterior probabilities |

The module analysis summarizes whether genes retain rhythmicity, gain or lose rhythmicity, change amplitude, shift phase, or show other rhythm-state remodeling patterns between two groups.

# Step 5: visualize rhythm-state modules

## Module heatmaps

The module heatmap summarizes expression patterns of genes assigned to rhythm-state modules. This tutorial shows all eight module IDs so users can inspect the full module structure in the rendered HTML page.

```{r module-heatmap-build}
hm_module <- plot_module_heatmap(
  count = count_use,
  res_phase = res_phase_use,
  gene_meta = res_module$module_result,
  metadata = metadata_use,
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

hm_module
```

```{r module-heatmap-0, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 0)
```

```{r module-heatmap-1, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 1)
```

```{r module-heatmap-2, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 2)
```

```{r module-heatmap-3, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 3)
```

```{r module-heatmap-4, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 4)
```

```{r module-heatmap-5, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 5)
```

```{r module-heatmap-6, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 6)
```

```{r module-heatmap-7, fig.width=7, fig.height=4.8, out.width="85%"}
plot(hm_module, module_id = 7)
```

The heatmaps are useful for checking whether genes assigned to each module show consistent phase-resolved expression patterns across the compared conditions.

## Module polar plot

The polar plot visualizes rhythmic remodeling in amplitude-phase space.

```{r module-polar-7, fig.width=6.5, fig.height=5.5, out.width="80%"}
p_polar <- plot_module_polar(
  gene_meta = res_module$module_result,
  module_id = 7,
  group1 = "A",
  group2 = "B"
)

p_polar$plot
```

This plot helps inspect how genes in a selected module are distributed by rhythmic amplitude and phase.

# Recommended minimal workflow

The minimal core workflow is:

```{r recommended-workflow, eval=FALSE}
data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta

res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

cell_use <- as.character(metadata$celltype) == "1"

res_param <- infer_gene_param(
  count = count[, cell_use, drop = FALSE],
  metadata = metadata[cell_use, , drop = FALSE],
  theta_pred = res_phase$theta_pred[cell_use],
  r_cell = res_phase$r_cell[cell_use],
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

hm_module <- plot_module_heatmap(
  count = count[, cell_use, drop = FALSE],
  res_phase = res_phase[cell_use, , drop = FALSE],
  gene_meta = res_module$module_result,
  metadata = metadata[cell_use, , drop = FALSE],
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

plot(hm_module, module_id = 0)
```

# Extension workflows

The following analyses are extensions of the core workflow. They are introduced here for orientation and described in separate tutorials.

## Extension 1: one-vs-rest rhythmic program discovery

The one-vs-rest workflow is designed for heterogeneous tissues or multi-cell-type datasets. It compares each target cell type against all remaining cell types.

```{r ovr-analysis, eval=FALSE}
res_ovr_param <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_ovr <- infer_ovr_rhythm(
  param_long = res_ovr_param$param_long,
  group_by = "celltype",
  q_cut = 0.05,
  amp_cut = 0.25,
  Verbose = FALSE
)

head(res_ovr$ovr_result)
```

## Extension 2: DTW-based phase-distribution similarity

The DTW workflow compares circular cell-phase distributions across celltype-by-condition combinations.

```{r dtw-analysis, eval=FALSE}
dtw_res <- run_dtw_similarity(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  min_cell = 10,
  Verbose = FALSE
)

head(dtw_res$histogram)
dtw_res$similarity
```

## Extension 3: circular phase-concentration rho analysis

The rho workflow summarizes how concentrated inferred cell phases are within cell types and conditions.

```{r rho-analysis, eval=FALSE}
rho_res <- run_phase_rho(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

rho_res
```

# Output summary

The core workflow produces three main result objects:

| Object | Description |
|---|---|
| `res_phase` | Cell-level inferred circadian phase and reliability statistics |
| `res_param` | Gene-level rhythmic parameter estimates |
| `res_module` | Differential rhythm-state module assignments |

The extension workflows produce additional result objects:

| Object | Description |
|---|---|
| `res_ovr` | One-vs-rest rhythmic program discovery results |
| `dtw_res` | DTW-based phase-distribution similarity results |
| `rho_res` | Group-level circular phase-concentration summaries |
| `rho_res_sample` | Sample-level circular phase-concentration summaries |

# Notes for real datasets

For real single-cell datasets, users should provide:

1. A gene-by-cell count matrix.
2. Cell-level metadata.
3. A sampling time column, such as `ZT`.
4. A stratum column for phase inference, such as cell type, sample, tissue region, or another biologically meaningful grouping variable.
5. A group or condition column for downstream comparison.

Recommended analysis strategy:

- infer cell phases globally or within appropriate strata;
- estimate gene-level rhythmic parameters within a biologically interpretable context;
- run module analysis for defined two-group comparisons;
- use one-vs-rest analysis for cell-type-specific rhythmic programs;
- use DTW and rho analyses for phase-distribution and phase-prioritization questions.

# Session information

```{r session-info}
sessionInfo()
```
