---
title: "Basic workflow: from cell phase inference to rhythm-state modules"
author: "Qian Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Basic workflow: from cell phase inference to rhythm-state modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This tutorial demonstrates the core workflow of `scRhythmBayes` using the built-in example dataset.

`scRhythmBayes` is designed for phase-resolved circadian analysis of single-cell transcriptomic data. The core workflow contains three main steps:

1. **Cell-level circadian phase inference**
   Infer latent circadian phase for each cell.

2. **Gene-level rhythmic parameter inference**
   Estimate gene-level rhythmic parameters, including mesor, mean expression, amplitude, phase, dropout rate, likelihood-ratio statistics, and FDR.

3. **Differential rhythm-state module analysis**
   Classify genes into interpretable rhythm-state remodeling modules between two biological groups.

This basic tutorial focuses on the main two-condition workflow. Extension analyses, including one-vs-rest rhythmic program discovery, DTW-based phase-distribution similarity, and circular phase-concentration rho analysis, are described in separate tutorials.

# Load package

```{r load-package}
library(scRhythmBayes)
```

# Load example data

`scRhythmBayes` includes an example dataset for demonstrating the full workflow.

```{r load-data}
data("srb_example_data")

names(srb_example_data)
```

The example object contains two components:

```{r data-components}
dim(srb_example_data$Count)
dim(srb_example_data$meta)
```

- `srb_example_data$Count`: a gene-by-cell sparse count matrix.
- `srb_example_data$meta`: cell-level metadata.

```{r show-metadata}
head(srb_example_data$meta)
```

The metadata contains:

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT` | Sampling Zeitgeber time |
| `celltype` | Cell-type label |
| `condition` | Condition or group label |
| `T_true` | Simulated true circadian time in hours |
| `Theta_true` | Simulated true circadian phase in radians |

`T_true` and `Theta_true` are included only for simulation-based validation. Real single-cell datasets usually do not contain these columns.

For convenience, define the input count matrix and metadata:

```{r define-inputs}
count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

# Step 1: infer cell-level circadian phase

The first step is to infer a latent circadian phase for each cell.

```{r infer-cell-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

The phase inference result contains one row per cell.

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT_input` | Observed sampling Zeitgeber time |
| `stratum` | Stratum used during phase inference |
| `theta_pred` | Inferred circadian phase in radians |
| `ZT_pred` | Inferred circadian phase converted to 24-hour time |
| `Rbar` | Aggregate phase concentration statistic |
| `r_cell` | Cell-level phase reliability weight |

The inferred phase can be used as the cell-level temporal coordinate for downstream rhythmic parameter estimation.

# Step 2: select one cell type for two-condition module analysis

The core rhythm-state module workflow is designed for two-group comparison within a defined biological context. In this example, we perform a condition comparison within one cell type.

```{r select-celltype}
cell_use <- as.character(metadata$celltype) == "1"

count_use <- count[, cell_use, drop = FALSE]
metadata_use <- metadata[cell_use, , drop = FALSE]
res_phase_use <- res_phase[cell_use, , drop = FALSE]

dim(count_use)
table(metadata_use$condition)
```

Here, cells from cell type `1` are used to compare condition `A` and condition `B`.

# Step 3: estimate gene-level rhythmic parameters

Next, estimate gene-level rhythmic parameters using the inferred cell phases.

```{r infer-gene-param}
res_param <- infer_gene_param(
  count = count_use,
  metadata = metadata_use,
  theta_pred = res_phase_use$theta_pred,
  r_cell = res_phase_use$r_cell,
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

The result includes long-format and wide-format parameter tables.

```{r show-param-long}
head(res_param$param_long)
```

```{r show-param-wide}
head(res_param$param_wide)
```

The long-format table stores gene-level rhythmic estimates for each group.

Common parameters include:

| Column | Description |
|---|---|
| `gene` | Gene name |
| `condition` | Group or condition |
| `Mesor` | Baseline rhythmic expression level |
| `Mean` | Mean expression level |
| `Amplitude` | Circadian oscillation amplitude |
| `Phase` | Gene-level rhythmic phase in radians |
| `Alpha` | Dispersion-related parameter |
| `Dropout_rate` | Estimated dropout rate |
| `LRT_stat` | Likelihood-ratio test statistic |
| `P_value` | Nominal rhythmicity test P value |
| `FDR` | Multiple-testing adjusted rhythmicity significance |

The wide-format table is used by the rhythm-state module analysis.

# Step 4: infer differential rhythm-state modules

The module analysis compares rhythmic parameters between two groups and assigns genes to interpretable rhythm-state remodeling classes.

```{r infer-module}
res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

head(res_module$module_result)
```

The module result contains gene-level group comparison statistics and posterior module assignments.

Important columns include:

| Column | Description |
|---|---|
| `gene` | Gene name |
| `group1`, `group2` | Compared groups |
| `FDR_A`, `FDR_B` | Rhythmicity FDR in each group |
| `Amplitude_A`, `Amplitude_B` | Rhythmic amplitude in each group |
| `Phase_A`, `Phase_B` | Rhythmic phase in each group |
| `delta_phi_rad` | Phase shift between groups in radians |
| `module_name` | Assigned rhythm-state remodeling module |
| `post_max` | Maximum posterior module probability |
| `margin` | Difference between top posterior probabilities |

The module analysis is intended to summarize whether genes remain rhythmic, gain or lose rhythmicity, shift phase, change amplitude, or show other rhythm-state remodeling patterns between conditions.

# Step 5: visualize rhythm-state modules with heatmap

The module heatmap summarizes expression patterns for genes assigned to rhythm-state modules.

```{r module-heatmap, eval=FALSE}
hm_module <- plot_module_heatmap(
  count = count_use,
  res_phase = res_phase_use,
  gene_meta = res_module$module_result,
  metadata = metadata_use,
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

hm_module
plot(hm_module, module_id = 0)
```

The heatmap object can be plotted by selecting a module ID. For example, `module_id = 0` displays genes assigned to module 0.

The exact module labels and interpretation depend on the module definitions used by the package.

# Step 6: visualize module-level rhythmic remodeling in polar space

The polar plot visualizes rhythmic remodeling in terms of amplitude and phase.

```{r module-polar, eval=FALSE}
p_polar <- plot_module_polar(
  gene_meta = res_module$module_result,
  module_id = 7,
  group1 = "A",
  group2 = "B"
)

p_polar$plot
```

This plot is useful for inspecting how genes in a selected module are distributed in rhythmic parameter space.

# Recommended core workflow

The minimal recommended workflow is:

```{r recommended-workflow, eval=FALSE}
data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta

res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

cell_use <- as.character(metadata$celltype) == "1"

res_param <- infer_gene_param(
  count = count[, cell_use, drop = FALSE],
  metadata = metadata[cell_use, , drop = FALSE],
  theta_pred = res_phase$theta_pred[cell_use],
  r_cell = res_phase$r_cell[cell_use],
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)
```

# Output summary

The core workflow produces three main result objects:

| Object | Description |
|---|---|
| `res_phase` | Cell-level inferred circadian phase and reliability statistics |
| `res_param` | Gene-level rhythmic parameter estimates |
| `res_module` | Differential rhythm-state module assignments |

# Notes for real datasets

For real single-cell datasets, users should provide:

1. A gene-by-cell count matrix.
2. Cell-level metadata.
3. A sampling time column, such as `ZT`.
4. A stratum column for phase inference, such as cell type, sample, tissue region, or another biologically meaningful grouping variable.
5. A group or condition column for downstream comparison.

The recommended analysis strategy is:

- infer cell phases globally or within appropriate strata;
- estimate gene-level rhythmic parameters within a biologically interpretable context;
- run module analysis for defined two-group comparisons;
- use extension workflows for multi-cell-type or phase-distribution questions.

# Session information

```{r session-info}
sessionInfo()
```
