---
title: "Rhythm-state module analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Rhythm-state module analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r module-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  out.width = "85%"
)
```

# Overview

This tutorial describes the rhythm-state module branch of `scRhythmBayes`.

The module branch is the main workflow for a **two-group rhythm-state comparison within a defined biological context**. A typical use case is condition `A` versus condition `B` within one selected cell type.

The workflow has five steps:

1. infer cell-level circadian phase;
2. select the biological context for comparison;
3. estimate gene-level rhythmic parameters for the two groups;
4. infer differential rhythm-state modules;
5. visualize module-level remodeling using heatmaps and polar plots.

This tutorial focuses on the condition comparison within cell type `1`. For direct cell-type comparisons or one-vs-rest discovery, see the basic workflow and OVR tutorials.

# Load package and example data

```{r module-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r module-data-summary}
dim(count)
head(metadata)
table(metadata$celltype, metadata$condition)
```

The required inputs are:

| Object | Description |
|---|---|
| `count` | Gene-by-cell count matrix |
| `metadata` | Cell-level metadata containing sampling time, cell type, and condition |
| `ZT` | Sampling Zeitgeber time column used for phase inference |
| `celltype` | Cell-type label used as the phase-inference stratum |
| `condition` | Two-group comparison variable for module analysis |

# Step 1: infer cell-level circadian phase

The module workflow starts from inferred cell phases. These phases are used as latent temporal coordinates for fitting gene-level rhythmic parameters.

```{r module-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

The output `res_phase` contains one row per cell.

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT_input` | Observed sampling Zeitgeber time |
| `stratum` | Stratum used during phase inference |
| `theta_pred` | Inferred cell-level circadian phase in radians |
| `ZT_pred` | Inferred phase converted to 24-hour time |
| `Rbar` | Aggregate phase concentration statistic |
| `r_cell` | Cell-level phase reliability weight |

The two key columns for downstream modeling are `theta_pred` and `r_cell`.

# Step 2: select one cell type for two-condition analysis

The recommended module-analysis design is to compare two groups within a defined biological context. Here, we select cell type `1` and compare condition `A` versus condition `B`.

```{r module-select}
cell_use <- as.character(metadata$celltype) == "1"

count_use <- count[, cell_use, drop = FALSE]
metadata_use <- metadata[cell_use, , drop = FALSE]
res_phase_use <- res_phase[cell_use, , drop = FALSE]

dim(count_use)
table(metadata_use$condition)
```

This subset contains only cells from cell type `1`. The downstream model estimates rhythmic parameters separately for condition `A` and condition `B`.

# Step 3: estimate gene-level rhythmic parameters

## Input

The gene-level model requires the selected count matrix, selected metadata, inferred cell phase, cell-level reliability weight, and a grouping variable.

| Argument | Description |
|---|---|
| `count` | Gene-by-cell count matrix for the selected cell subset |
| `metadata` | Metadata for the same cells |
| `theta_pred` | Inferred cell phase from `res_phase` |
| `r_cell` | Cell-level phase reliability weight |
| `strata_by` | Grouping variable for gene-level parameter estimation, here `condition` |

## Run gene-level parameter inference

```{r module-param}
res_param <- infer_gene_param(
  count = count_use,
  metadata = metadata_use,
  theta_pred = res_phase_use$theta_pred,
  r_cell = res_phase_use$r_cell,
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

## Inspect output tables

```{r module-param-long}
head(res_param$param_long)
```

```{r module-param-wide}
head(res_param$param_wide)
```

The long-format table stores rhythmic parameters for each group. The wide-format table is used for paired module comparison.

| Output | Description |
|---|---|
| `param_long` | Long-format gene-level rhythmic parameter estimates for each group |
| `param_wide` | Wide-format paired parameter table for downstream module analysis |
| `bootstrap_res` | Bootstrap results if available |

Common columns include:

| Column | Meaning |
|---|---|
| `gene` | Gene name |
| `Mesor` | Baseline rhythmic expression level |
| `Mean` | Mean expression level |
| `Amplitude` | Circadian oscillation amplitude |
| `Phase` | Rhythmic phase in radians |
| `Alpha` | Dispersion-related parameter |
| `Dropout_rate` | Estimated dropout rate |
| `LRT_stat` | Likelihood-ratio test statistic |
| `P_value` | Nominal rhythmicity test P value |
| `FDR` | Adjusted rhythmicity significance |

# Step 4: infer differential rhythm-state modules

The module model compares rhythmic parameters between two groups and assigns genes to interpretable rhythm-state remodeling classes.

```{r module-infer}
res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

head(res_module$module_result)
```

```{r module-counts}
table(res_module$module_result$module_name, useNA = "ifany")
```

The module result summarizes whether genes remain rhythmic, gain rhythmicity, lose rhythmicity, shift phase, change amplitude, or show other rhythm-state remodeling patterns.

Important output columns include:

| Column | Description |
|---|---|
| `gene` | Gene name |
| `compare_by` | Comparison variable |
| `group1`, `group2` | Compared groups |
| `FDR_A`, `FDR_B` | Rhythmicity FDR in each group |
| `Amplitude_A`, `Amplitude_B` | Rhythmic amplitude in each group |
| `Phase_A`, `Phase_B` | Rhythmic phase in each group |
| `delta_phi_rad` | Phase shift between groups in radians |
| `module_name` | Assigned rhythm-state remodeling module |
| `post_max` | Maximum posterior module probability |
| `margin` | Difference between top posterior probabilities |

# Step 5: visualize rhythm-state modules

## Build the module heatmap object

The module heatmap summarizes phase-ordered expression patterns for genes assigned to rhythm-state modules.

```{r module-heatmap-build}
hm_module <- plot_module_heatmap(
  count = count_use,
  res_phase = res_phase_use,
  gene_meta = res_module$module_result,
  metadata = metadata_use,
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

hm_module
```

## Plot selected module heatmaps

The following plots display selected module IDs. This keeps the tutorial compact while still showing the heatmap output format.

```{r module-heatmap-0, fig.width=7, fig.height=5}
plot(hm_module, module_id = 0)
```

```{r module-heatmap-1, fig.width=7, fig.height=5}
plot(hm_module, module_id = 1)
```

```{r module-heatmap-7, fig.width=7, fig.height=5}
plot(hm_module, module_id = 7)
```

The heatmap shows expression patterns ordered by inferred cell phase and separated by condition. It is useful for checking whether genes assigned to a module show coherent phase-resolved expression patterns across the compared groups.

## Optional: plot all eight module heatmaps

If a complete visual inspection is needed, all eight module IDs can be plotted.

```{r module-heatmap-all, eval=FALSE}
plot(hm_module, module_id = 0)
plot(hm_module, module_id = 1)
plot(hm_module, module_id = 2)
plot(hm_module, module_id = 3)
plot(hm_module, module_id = 4)
plot(hm_module, module_id = 5)
plot(hm_module, module_id = 6)
plot(hm_module, module_id = 7)
```

Keeping the full set as `eval = FALSE` can reduce vignette build time and avoid generating an overly long article page. If the goal is a fully visual GitHub Pages tutorial, these lines can be split into separate evaluated chunks.

# Step 6: plot rhythm remodeling in polar space

The polar plot displays rhythmic remodeling in amplitude-phase space. Phase is represented by angular position and amplitude by radial distance.

```{r module-polar, fig.width=6.5, fig.height=5.5}
p_polar <- plot_module_polar(
  gene_meta = res_module$module_result,
  module_id = 7,
  group1 = "A",
  group2 = "B"
)

p_polar$plot
```

This plot is useful for inspecting how genes in a selected module are distributed by rhythmic amplitude and phase.

# Recommended minimal workflow

```{r module-minimal, eval=FALSE}
data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta

res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

cell_use <- as.character(metadata$celltype) == "1"

count_use <- count[, cell_use, drop = FALSE]
metadata_use <- metadata[cell_use, , drop = FALSE]
res_phase_use <- res_phase[cell_use, , drop = FALSE]

res_param <- infer_gene_param(
  count = count_use,
  metadata = metadata_use,
  theta_pred = res_phase_use$theta_pred,
  r_cell = res_phase_use$r_cell,
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

hm_module <- plot_module_heatmap(
  count = count_use,
  res_phase = res_phase_use,
  gene_meta = res_module$module_result,
  metadata = metadata_use,
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

plot(hm_module, module_id = 0)
```

# Notes for real datasets

For real datasets, the module branch should be used when the comparison is clearly defined as two groups within a biological context.

Typical designs include:

| Biological question | Suggested setup |
|---|---|
| Disease versus control within one cell type | Subset one cell type and set `strata_by = "condition"` |
| Treatment versus baseline within one cell type | Subset one cell type and set `strata_by = "condition"` |
| Region A versus region B within one cell type | Subset one cell type and set `strata_by` to the region column |
| Two selected cell types | Estimate parameters by `celltype`, then compare two selected labels |

For broad cell-type-specific discovery across many cell types, the OVR workflow is usually more appropriate than repeatedly running pairwise module comparisons.

# Session information

```{r module-session}
sessionInfo()
```
