---
title: "Rhythm-state module analysis"
author: "Qian Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Rhythm-state module analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r module-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```


# Overview

This tutorial describes the rhythm-state module branch of `scRhythmBayes`.

This branch is the main workflow for a **two-group rhythm-state comparison within a defined biological context**, for example condition `A` versus condition `B` within one cell type.

The workflow is:

1. infer cell-level circadian phase;
2. select a biological context for comparison;
3. estimate gene-level rhythmic parameters for the two groups;
4. infer differential rhythm-state modules;
5. visualize module-level remodeling using heatmaps and polar plots.

# Load package and example data

```{r module-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r module-data-summary}
dim(count)
head(metadata)
table(metadata$celltype, metadata$condition)
```

# Step 1: infer cell-level circadian phase

```{r module-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

The output `res_phase` contains one row per cell. The key columns are `theta_pred`, the inferred phase in radians, and `r_cell`, the cell-level phase reliability weight.

# Step 2: select one cell type for two-condition analysis

```{r module-select}
cell_use <- as.character(metadata$celltype) == "1"

count_use <- count[, cell_use, drop = FALSE]
metadata_use <- metadata[cell_use, , drop = FALSE]
res_phase_use <- res_phase[cell_use, , drop = FALSE]

dim(count_use)
table(metadata_use$condition)
```

This example compares condition `A` and condition `B` within cell type `1`.

# Step 3: estimate gene-level rhythmic parameters

```{r module-param}
res_param <- infer_gene_param(
  count = count_use,
  metadata = metadata_use,
  theta_pred = res_phase_use$theta_pred,
  r_cell = res_phase_use$r_cell,
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

```{r module-param-long}
head(res_param$param_long)
```

```{r module-param-wide}
head(res_param$param_wide)
```

The long-format table stores rhythmic parameters for each group. The wide-format table is used for paired module comparison.

| Column | Meaning |
|---|---|
| `Mesor` | Baseline rhythmic expression level |
| `Mean` | Mean expression level |
| `Amplitude` | Circadian oscillation amplitude |
| `Phase` | Rhythmic phase in radians |
| `FDR` | Adjusted rhythmicity significance |

# Step 4: infer differential rhythm-state modules

```{r module-infer}
res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)

head(res_module$module_result)
```

```{r module-counts}
table(res_module$module_result$module_name, useNA = "ifany")
```

The module result summarizes whether genes remain rhythmic, gain rhythmicity, lose rhythmicity, shift phase, change amplitude, or show other rhythm-state remodeling patterns.

# Step 5: plot the module heatmap

```{r module-heatmap}
hm_module <- plot_module_heatmap(
  count = count_use,
  res_phase = res_phase_use,
  gene_meta = res_module$module_result,
  metadata = metadata_use,
  group_by = "condition",
  group1 = "A",
  group2 = "B"
)

hm_module
plot(hm_module, module_id = 0)
```

The heatmap shows expression patterns ordered by inferred cell phase and separated by condition.

# Step 6: plot rhythm remodeling in polar space

```{r module-polar}
p_polar <- plot_module_polar(
  gene_meta = res_module$module_result,
  module_id = 7,
  group1 = "A",
  group2 = "B"
)

p_polar$plot
```

The polar plot displays rhythmic remodeling in amplitude-phase space. Phase is represented by angular position and amplitude by radial distance.

# Recommended minimal workflow

```{r module-minimal, eval=FALSE}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

cell_use <- as.character(metadata$celltype) == "1"

res_param <- infer_gene_param(
  count = count[, cell_use, drop = FALSE],
  metadata = metadata[cell_use, , drop = FALSE],
  theta_pred = res_phase$theta_pred[cell_use],
  r_cell = res_phase$r_cell[cell_use],
  strata_by = "condition",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_module <- infer_rhythm_module(
  param_wide = res_param$param_wide,
  compare_by = "condition",
  group1 = "A",
  group2 = "B",
  out_dir = NULL,
  Verbose = FALSE
)
```

# Session information

```{r module-session}
sessionInfo()
```
