---
title: "One-vs-rest rhythmic program discovery"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{One-vs-rest rhythmic program discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ovr-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  fig.align = "center",
  out.width = "85%"
)
```

# Overview

This tutorial describes the one-vs-rest rhythmic program discovery workflow in `scRhythmBayes`.

The one-vs-rest branch is designed for heterogeneous tissues and multi-cell-type datasets. It compares one target cell type against all remaining cell types and identifies cell-type-specific rhythmic programs.

This branch starts from the gene-level rhythmic parameter layer. It is different from pairwise rhythm-state module analysis:

| Analysis | Main question | Typical use case |
|---|---|---|
| Pairwise module analysis | How do rhythmic states differ between two predefined groups? | Condition `A` versus condition `B` within one cell type |
| One-vs-rest analysis | Which rhythmic programs are specific to one target cell type relative to all other cell types? | Cell-type-specific rhythmic program discovery in heterogeneous tissue |

In this tutorial, the workflow is:

1. infer cell-level circadian phase;
2. estimate gene-level rhythmic parameters by cell type;
3. compare each target cell type against the pooled rest;
4. inspect the OVR output table;
5. visualize cell-type-associated rhythmic programs using the OVR heatmap;
6. extract target-specific candidate genes.

# Load package and example data

```{r ovr-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r ovr-data-summary}
dim(count)
head(metadata)
table(metadata$celltype)
table(metadata$celltype, metadata$condition)
```

The two required input objects are:

| Object | Description |
|---|---|
| `count` | Gene-by-cell count matrix |
| `metadata` | Cell-level metadata containing cell IDs, cell-type labels, and sampling time |

For the example dataset, `metadata$celltype` defines the cell-type groups used for OVR testing.

# Step 1: infer cell-level circadian phase

The OVR workflow first requires a cell-level circadian phase estimate. Here, phase is inferred using `celltype` as the stratum.

```{r ovr-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

The key output columns used downstream are:

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `theta_pred` | Inferred cell-level circadian phase in radians |
| `ZT_pred` | Inferred cell-level circadian phase converted to 24-hour time |
| `r_cell` | Cell-level phase reliability weight |

# Step 2: estimate gene-level rhythmic parameters by cell type

For OVR analysis, gene-level rhythmic parameters should be estimated across cell types. Therefore, use `strata_by = "celltype"`.

```{r ovr-param}
res_ovr_param <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

The main table used by `infer_ovr_rhythm()` is `param_long`.

```{r ovr-param-long}
head(res_ovr_param$param_long)
```

The long-format parameter table contains one row per gene and cell type.

| Column | Description |
|---|---|
| `gene` | Gene name |
| `celltype` | Cell-type group |
| `Mesor` | Baseline rhythmic expression level |
| `Mean` | Mean expression level |
| `Amplitude` | Rhythmic amplitude |
| `Phase` | Rhythmic phase in radians |
| `FDR` | Adjusted rhythmicity significance |

# Step 3: run one-vs-rest analysis

`infer_ovr_rhythm()` compares each target cell type against the pooled rest.

```{r ovr-infer}
res_ovr <- infer_ovr_rhythm(
  param_long = res_ovr_param$param_long,
  group_by = "celltype",
  q_cut = 0.05,
  amp_cut = 0.25,
  Verbose = FALSE
)

head(res_ovr$ovr_result)
```

The main thresholds are:

| Argument | Description |
|---|---|
| `q_cut` | FDR cutoff for rhythmicity |
| `amp_cut` | Minimum target-group amplitude cutoff used for target-specific calls |

```{r ovr-summary}
table(res_ovr$ovr_result$rhythm_pattern, useNA = "ifany")
table(res_ovr$ovr_result$target_specific_flag, useNA = "ifany")
```

# Interpret the OVR result

The OVR result is a gene-by-target-cell-type table.

| Column | Meaning |
|---|---|
| `gene` | Gene name |
| `TargetCellType` | Target cell type being tested |
| `RestCellTypes` | Pooled non-target cell types |
| `rhythm_pattern` | One-vs-rest rhythm-state pattern |
| `FDR_group1` | Rhythmicity FDR in the target group |
| `FDR_group2` | Rhythmicity FDR in the pooled rest group |
| `Mean_group1`, `Mean_group2` | Mean expression in target and rest groups |
| `Amplitude_group1`, `Amplitude_group2` | Rhythmic amplitude in target and rest groups |
| `Phase_group1`, `Phase_group2` | Rhythmic phase in target and rest groups |
| `Delta_Mean` | Target-minus-rest mean difference |
| `Delta_Amp` | Target-minus-rest amplitude difference |
| `target_specific_flag` | Whether the gene is classified as target-specific |

A typical target-specific rhythmic program is expected to have strong rhythmic evidence and sufficient amplitude in the target cell type, with weaker or absent rhythmic evidence in the pooled rest.

# Step 4: plot the OVR heatmap

The OVR heatmap visualizes target-associated rhythmic programs across the heterogeneous dataset.

```{r ovr-heatmap, fig.width=7.5, fig.height=6}
ht_ovr <- plot_ovr_heatmap(
  count = count,
  res_phase = res_phase,
  ovr_result = res_ovr$ovr_result,
  metadata = metadata,
  celltype_by = "celltype",
  Verbose = FALSE
)

ht_ovr
plot(ht_ovr)
```

The heatmap is intended to help inspect whether OVR-selected genes show cell-type-associated phase-resolved expression patterns.

# Step 5: extract target-specific candidates

The `target_specific_flag` column can be used to extract candidate cell-type-specific rhythmic genes.

```{r ovr-candidates}
target_specific <- res_ovr$ovr_result[
  res_ovr$ovr_result$target_specific_flag %in% TRUE,
]

head(target_specific)
```

The number of target-specific genes depends on signal strength, the selected thresholds, and the structure of the dataset.

```{r ovr-candidates-by-celltype}
if (nrow(target_specific) > 0) {
  table(target_specific$TargetCellType)
}
```

If no target-specific genes are detected in a small example dataset, this does not indicate function failure. In real datasets, users may adjust `q_cut` and `amp_cut` based on the expected signal and desired stringency.

# Recommended minimal workflow

```{r ovr-minimal, eval=FALSE}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

res_ovr_param <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_ovr <- infer_ovr_rhythm(
  param_long = res_ovr_param$param_long,
  group_by = "celltype",
  q_cut = 0.05,
  amp_cut = 0.25,
  Verbose = FALSE
)

ht_ovr <- plot_ovr_heatmap(
  count = count,
  res_phase = res_phase,
  ovr_result = res_ovr$ovr_result,
  metadata = metadata,
  celltype_by = "celltype",
  Verbose = FALSE
)

plot(ht_ovr)
```

# Notes for real datasets

For real datasets, the OVR workflow is most useful when:

1. the dataset contains multiple biologically meaningful cell types;
2. cell types have sufficient cell numbers for rhythmic parameter estimation;
3. the user wants cell-type-specific rhythmic programs rather than a predefined pairwise condition comparison;
4. the target cell type may have rhythmic genes that are absent, weaker, or phase-shifted in the rest of the tissue.

For direct comparison of two selected groups, use `infer_rhythm_module()` instead. For target-versus-all discovery across many cell types, use `infer_ovr_rhythm()`.

# Session information

```{r ovr-session}
sessionInfo()
```
