---
title: "One-vs-rest rhythmic program discovery"
author: "Qian Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{One-vs-rest rhythmic program discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r ovr-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```


# Overview

This tutorial describes one-vs-rest rhythmic program discovery in `scRhythmBayes`.

The one-vs-rest branch is designed for heterogeneous tissues and multi-cell-type datasets. It compares one target cell type against all remaining cell types and identifies cell-type-specific rhythmic programs.

This branch is an extension from the gene-level rhythmic parameter layer. It is different from module analysis: module analysis is for a predefined two-condition comparison, while OVR is for cell-type-specific discovery.

# Load package and example data

```{r ovr-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r ovr-data-summary}
dim(count)
head(metadata)
table(metadata$celltype)
```

# Step 1: infer cell-level circadian phase

```{r ovr-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

# Step 2: estimate gene-level parameters by cell type

```{r ovr-param}
res_ovr_param <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)
```

```{r ovr-param-long}
head(res_ovr_param$param_long)
```

# Step 3: run one-vs-rest analysis

```{r ovr-infer}
res_ovr <- infer_ovr_rhythm(
  param_long = res_ovr_param$param_long,
  group_by = "celltype",
  q_cut = 0.05,
  amp_cut = 0.25,
  Verbose = FALSE
)

head(res_ovr$ovr_result)
```

```{r ovr-summary}
table(res_ovr$ovr_result$rhythm_pattern, useNA = "ifany")
table(res_ovr$ovr_result$target_specific_flag, useNA = "ifany")
```

# Interpret the OVR result

| Column | Meaning |
|---|---|
| `gene` | Gene name |
| `TargetCellType` | Target cell type being tested |
| `RestCellTypes` | Pooled non-target cell types |
| `rhythm_pattern` | One-vs-rest rhythm-state pattern |
| `FDR_group1` | Rhythmicity FDR in the target group |
| `FDR_group2` | Rhythmicity FDR in the pooled rest group |
| `Mean_group1`, `Mean_group2` | Mean expression in target and rest |
| `Amplitude_group1`, `Amplitude_group2` | Rhythmic amplitude in target and rest |
| `Phase_group1`, `Phase_group2` | Rhythmic phase in target and rest |
| `Delta_Mean` | Target-minus-rest mean difference |
| `Delta_Amp` | Target-minus-rest amplitude difference |
| `target_specific_flag` | Whether the gene is classified as target-specific |

# Step 4: plot the OVR heatmap

```{r ovr-heatmap}
ht_ovr <- plot_ovr_heatmap(
  count = count,
  res_phase = res_phase,
  ovr_result = res_ovr$ovr_result,
  metadata = metadata,
  celltype_by = "celltype",
  Verbose = FALSE
)

ht_ovr
plot(ht_ovr)
```

The heatmap visualizes cell-type-associated rhythmic programs across the heterogeneous dataset.

# Step 5: extract target-specific candidates

```{r ovr-candidates}
target_specific <- res_ovr$ovr_result[
  res_ovr$ovr_result$target_specific_flag %in% TRUE,
]

head(target_specific)
```

If no target-specific genes are detected in a small example dataset, this does not indicate function failure. The result depends on the signal strength, the filtering thresholds, and the simulated structure of the example data.

# Recommended minimal workflow

```{r ovr-minimal, eval=FALSE}
res_ovr_param <- infer_gene_param(
  count = count,
  metadata = metadata,
  theta_pred = res_phase$theta_pred,
  r_cell = res_phase$r_cell,
  strata_by = "celltype",
  B = 10,
  n_worker = 1,
  setseed = 1,
  Verbose = FALSE
)

res_ovr <- infer_ovr_rhythm(
  param_long = res_ovr_param$param_long,
  group_by = "celltype",
  q_cut = 0.05,
  amp_cut = 0.25,
  Verbose = FALSE
)

ht_ovr <- plot_ovr_heatmap(
  count = count,
  res_phase = res_phase,
  ovr_result = res_ovr$ovr_result,
  metadata = metadata,
  celltype_by = "celltype",
  Verbose = FALSE
)
```

# Session information

```{r ovr-session}
sessionInfo()
```
