---
title: "DTW and rho phase-prioritization analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{DTW and rho phase-prioritization analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r phaseprior-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  out.width = "85%",
  fig.align = "center"
)
```

# Overview

This tutorial describes two phase-level extension analyses in `scRhythmBayes`:

1. **DTW-based phase-distribution similarity**
2. **Circular phase-concentration rho analysis**

These analyses are useful for prioritizing cell types, conditions, or celltype-by-condition combinations that show differences in inferred circadian phase distributions.

Unlike rhythm-state module analysis or one-vs-rest rhythmic program discovery, DTW and rho analyses operate mainly at the **cell-phase level**. They use the inferred cell-level phase result and cell-level metadata, rather than gene-level parameter tables.

The typical use case is:

1. infer cell-level circadian phase;
2. compare phase distributions across cell types and groups;
3. prioritize cell types or conditions with strong phase-distribution remodeling;
4. perform downstream gene-level rhythm-state analysis on selected biological contexts.

# Load package and example data

```{r phaseprior-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r phaseprior-data-summary}
dim(count)
head(metadata)
table(metadata$celltype, metadata$condition)
```

The example dataset contains:

| Object | Description |
|---|---|
| `count` | Gene-by-cell sparse count matrix |
| `metadata` | Cell-level metadata containing sampling time, cell type, condition, and simulated true phase |

The key metadata columns used in this tutorial are:

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `ZT` | Sampling Zeitgeber time |
| `celltype` | Cell-type label |
| `condition` | Group or condition label |
| `T_true` | Simulated true circadian time in hours |
| `Theta_true` | Simulated true circadian phase in radians |

# Step 1: infer cell-level circadian phase

Both DTW and rho analyses start from the inferred cell-level circadian phase.

```{r phaseprior-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

The main columns used by downstream phase-prioritization analyses are:

| Column | Description |
|---|---|
| `cell` | Cell identifier |
| `theta_pred` | Inferred cell-level circadian phase in radians |
| `ZT_pred` | Inferred phase converted to 24-hour time |
| `r_cell` | Cell-level phase reliability weight |

# Analysis design

The phase-prioritization branch can be used in several common settings.

| Analysis goal | Function | Main grouping variables | Main output |
|---|---|---|---|
| Compare full phase-distribution shapes | `run_dtw_similarity()` | `celltype_by`, `group_by` | DTW distance and similarity matrices |
| Summarize phase concentration by group | `run_phase_rho()` | `celltype_by`, `group_by` | Group-level rho summary |
| Compare sample-level phase concentration | `run_phase_rho()` with `sample_by` | `celltype_by`, `group_by`, `sample_by` | Sample-level rho table and group comparison |

DTW is useful when the question is whether two phase distributions have similar shapes. Rho is useful when the question is whether cells within a cell type or group are concentrated around a phase or broadly dispersed.

# Part I: DTW-based phase-distribution similarity

## Input

The DTW analysis requires:

| Argument | Description |
|---|---|
| `res_phase` | Cell-level phase inference result from `infer_cell_phase()` |
| `metadata` | Cell-level metadata |
| `celltype_by` | Metadata column defining cell type or biological compartment |
| `group_by` | Metadata column defining condition or group |
| `min_cell` | Minimum number of cells required for each celltype-by-group combination |

The function constructs circular phase histograms for each celltype-by-group combination and compares these histograms using dynamic time warping.

## Run DTW similarity analysis

```{r dtw-run}
dtw_res <- run_dtw_similarity(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  min_cell = 10,
  Verbose = FALSE
)

dtw_res
```

## Inspect circular phase histograms

```{r dtw-histogram}
head(dtw_res$histogram)
```

The histogram table contains binned circular phase distributions for each celltype-by-condition combination.

| Column | Description |
|---|---|
| `combo` | Combined celltype-by-group label |
| `cell_type` | Cell-type label |
| `Group` | Group or condition label |
| `bin` | Circular phase bin index |
| `bin_start`, `bin_end` | Phase-bin boundaries in radians |
| `bin_mid` | Phase-bin midpoint |
| `prop` | Proportion of cells in the phase bin |

## Inspect DTW similarity matrix

```{r dtw-similarity}
dtw_res$similarity
```

The result object contains:

| Output | Meaning |
|---|---|
| `data` | Merged cell-level phase and metadata table |
| `histogram` | Circular phase histogram for each celltype-by-group combination |
| `distance` | Pairwise DTW distance matrix |
| `similarity` | Scaled 0-to-1 DTW similarity matrix |
| `annotation` | Annotation table for heatmap rows and columns |

A higher similarity value indicates a more similar inferred phase-distribution profile. A lower similarity value indicates a stronger phase-distribution difference.

## Plot DTW heatmap

```{r dtw-heatmap, fig.width=7, fig.height=5, out.width="85%", fig.align="center"}
ht_dtw <- plot_dtw_heatmap(
  dtw_res = dtw_res
)

ht_dtw
plot(ht_dtw)
```

The DTW heatmap provides a compact overview of phase-distribution similarity across celltype-by-condition combinations.

# Part II: circular phase-concentration rho analysis

## Concept

The rho analysis summarizes the concentration of inferred cell phases. Larger rho indicates that cells are more concentrated around a common phase. Smaller rho indicates a more dispersed phase distribution.

This analysis is useful for identifying cell types or conditions with strong phase alignment or phase dispersion.

## Input

The group-level rho analysis requires:

| Argument | Description |
|---|---|
| `res_phase` | Cell-level phase inference result |
| `metadata` | Cell-level metadata |
| `celltype_by` | Metadata column defining cell type |
| `group_by` | Metadata column defining condition or group |
| `theta_by` | Column in `res_phase` containing inferred phase in radians |
| `group_levels` | Optional display or comparison order for groups |
| `min_cell` | Minimum number of cells required per celltype-by-group combination |

## Run group-level rho analysis

```{r rho-group-run}
rho_res <- run_phase_rho(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

rho_res
```

## Inspect group-level rho table

```{r rho-group-table}
rho_res$rho_group
```

The group-level rho table contains:

| Column | Description |
|---|---|
| `cell_type` | Cell-type label |
| `Group` | Group or condition label |
| `n_cell` | Number of cells used |
| `rho` | Circular phase concentration statistic |
| `mean_phase_rad` | Mean circular phase in radians |
| `mean_phase_hour` | Mean circular phase converted to 24-hour time |

## Plot group-level rho

```{r rho-group-plot, fig.width=7, fig.height=5, out.width="85%", fig.align="center"}
p_rho <- plot_phase_rho(rho_res)

p_rho$plot
```

The plot summarizes phase concentration across cell types and groups.

# Part III: sample-level rho analysis

Sample-level rho analysis is useful when the dataset contains biological sample IDs, replicate IDs, patient IDs, animal IDs, or other sample-level units.

The example dataset does not contain real biological sample IDs. The following code creates artificial sample labels for demonstration only.

```{r rho-create-sample}
set.seed(1)

meta_rho_test <- metadata

if (!"cell" %in% colnames(meta_rho_test)) {
  meta_rho_test$cell <- rownames(meta_rho_test)
}

meta_rho_test$Sample <- NA_character_

sample_per_group <- 4

split_key <- interaction(
  meta_rho_test$celltype,
  meta_rho_test$condition,
  drop = TRUE
)

for (kk in levels(split_key)) {
  idx <- which(split_key == kk)

  meta_rho_test$Sample[idx] <- sample(
    paste0("S", seq_len(sample_per_group)),
    size = length(idx),
    replace = TRUE
  )
}

head(meta_rho_test)
```

## Run sample-level rho analysis

```{r rho-sample-run}
rho_res_sample <- run_phase_rho(
  res_phase = res_phase,
  metadata = meta_rho_test,
  celltype_by = "celltype",
  group_by = "condition",
  sample_by = "Sample",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

head(rho_res_sample$rho_sample)
```

The sample-level rho table contains one row per celltype-by-group-by-sample unit.

| Column | Description |
|---|---|
| `cell_type` | Cell-type label |
| `Group` | Group or condition label |
| `Sample` | Sample identifier |
| `n_cell` | Number of cells in the sample-level unit |
| `rho` | Sample-level circular phase concentration |
| `mean_phase_rad` | Mean circular phase in radians |
| `mean_phase_hour` | Mean circular phase converted to 24-hour time |

## Plot sample-level rho

```{r rho-sample-plot, fig.width=7, fig.height=5, out.width="85%", fig.align="center"}
p_rho_sample <- plot_phase_rho(rho_res_sample)

p_rho_sample$plot
```

The sample-level plot is more appropriate for real biological datasets with replicate structure. It can help distinguish cell-level phase-distribution patterns from sample-level variability.

# Recommended minimal workflow

```{r phaseprior-minimal, eval=FALSE}
dtw_res <- run_dtw_similarity(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  min_cell = 10,
  Verbose = FALSE
)

ht_dtw <- plot_dtw_heatmap(dtw_res = dtw_res)
plot(ht_dtw)

rho_res <- run_phase_rho(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

p_rho <- plot_phase_rho(rho_res)
p_rho$plot
```

# Output summary

| Object | Description |
|---|---|
| `dtw_res` | DTW-based phase-distribution similarity result |
| `dtw_res$histogram` | Circular phase histograms by celltype-by-group combination |
| `dtw_res$distance` | Pairwise DTW distance matrix |
| `dtw_res$similarity` | Scaled DTW similarity matrix |
| `rho_res` | Group-level circular phase-concentration result |
| `rho_res$rho_group` | Group-level rho summary |
| `rho_res_sample` | Sample-level circular phase-concentration result |
| `rho_res_sample$rho_sample` | Sample-level rho summary |

# Notes for real datasets

For real datasets, users should choose grouping variables based on the biological question.

Use DTW analysis when the main question is:

- Which cell types or conditions have similar inferred phase distributions?
- Which celltype-by-condition combinations show phase-distribution remodeling?
- Which biological contexts should be prioritized for downstream rhythm-state module analysis?

Use rho analysis when the main question is:

- Are cells in a given cell type or condition phase-concentrated?
- Does phase concentration differ between conditions?
- Is phase concentration consistent across biological samples?

For sample-level rho analysis, `sample_by` should correspond to real biological replicate structure, such as animal ID, patient ID, donor ID, or sample ID.

# Session information

```{r phaseprior-session}
sessionInfo()
```
