---
title: "DTW and rho phase-prioritization analysis"
author: "Qian Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{DTW and rho phase-prioritization analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r phaseprior-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```


# Overview

This tutorial describes two phase-level extension analyses:

1. DTW-based phase-distribution similarity;
2. circular phase-concentration rho analysis.

These analyses are useful for prioritizing cell types or conditions that show phase-distribution remodeling before downstream gene-level modeling.

# Load package and example data

```{r phaseprior-load}
library(scRhythmBayes)

data("srb_example_data")

count <- srb_example_data$Count
metadata <- srb_example_data$meta
```

```{r phaseprior-data-summary}
dim(count)
head(metadata)
table(metadata$celltype, metadata$condition)
```

# Step 1: infer cell-level circadian phase

```{r phaseprior-phase}
res_phase <- infer_cell_phase(
  count = count,
  metadata = metadata,
  ZT_by = "ZT",
  strata_by = "celltype",
  Species = "mouse",
  setseed = 1,
  Verbose = FALSE
)

head(res_phase)
```

# Part I: DTW-based phase-distribution similarity

## Run DTW similarity analysis

```{r dtw-run}
dtw_res <- run_dtw_similarity(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  min_cell = 10,
  Verbose = FALSE
)

dtw_res
```

```{r dtw-histogram}
head(dtw_res$histogram)
```

```{r dtw-similarity}
dtw_res$similarity
```

| Output | Meaning |
|---|---|
| `data` | Merged cell-level phase and metadata table |
| `histogram` | Circular phase histogram for each celltype-by-group combination |
| `distance` | Pairwise DTW distance matrix |
| `similarity` | Scaled 0-to-1 DTW similarity matrix |
| `annotation` | Annotation table for heatmap rows and columns |

A higher similarity value indicates a more similar inferred phase-distribution profile.

## Plot DTW heatmap

```{r dtw-heatmap}
ht_dtw <- plot_dtw_heatmap(
  dtw_res = dtw_res
)

ht_dtw
plot(ht_dtw)
```

# Part II: circular phase-concentration rho analysis

## Run group-level rho analysis

```{r rho-group-run}
rho_res <- run_phase_rho(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

rho_res
```

```{r rho-group-table}
rho_res$rho_group
```

## Plot group-level rho

```{r rho-group-plot}
p_rho <- plot_phase_rho(rho_res)

p_rho$plot
```

A larger rho indicates stronger phase concentration. A smaller rho indicates a more dispersed phase distribution.

# Part III: sample-level rho analysis

The example dataset does not contain real biological sample IDs. The following code creates artificial sample labels for demonstration only.

```{r rho-create-sample}
set.seed(1)

meta_rho_test <- metadata

if (!"cell" %in% colnames(meta_rho_test)) {
  meta_rho_test$cell <- rownames(meta_rho_test)
}

meta_rho_test$Sample <- NA_character_

sample_per_group <- 4

split_key <- interaction(
  meta_rho_test$celltype,
  meta_rho_test$condition,
  drop = TRUE
)

for (kk in levels(split_key)) {
  idx <- which(split_key == kk)

  meta_rho_test$Sample[idx] <- sample(
    paste0("S", seq_len(sample_per_group)),
    size = length(idx),
    replace = TRUE
  )
}

head(meta_rho_test)
```

## Run sample-level rho analysis

```{r rho-sample-run}
rho_res_sample <- run_phase_rho(
  res_phase = res_phase,
  metadata = meta_rho_test,
  celltype_by = "celltype",
  group_by = "condition",
  sample_by = "Sample",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

head(rho_res_sample$rho_sample)
```

## Plot sample-level rho

```{r rho-sample-plot}
p_rho_sample <- plot_phase_rho(rho_res_sample)

p_rho_sample$plot
```

# Recommended minimal workflow

```{r phaseprior-minimal, eval=FALSE}
dtw_res <- run_dtw_similarity(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  min_cell = 10,
  Verbose = FALSE
)

ht_dtw <- plot_dtw_heatmap(dtw_res = dtw_res)

rho_res <- run_phase_rho(
  res_phase = res_phase,
  metadata = metadata,
  celltype_by = "celltype",
  group_by = "condition",
  theta_by = "theta_pred",
  group_levels = c("A", "B"),
  min_cell = 10,
  Verbose = FALSE
)

p_rho <- plot_phase_rho(rho_res)
```

# Session information

```{r phaseprior-session}
sessionInfo()
```
